---
title: "police_shooting"
output:
  html_document:
    df_print: paged
date: "2023-03-05"
---


---

## *Reading in data*
```{r}
df <- read.csv("NYPD_Shooting_Incident_Data__Historic_.csv")
head(df)
```

---

## *Tidying up the data*

I'll remove extraneous columns, like all of the latitude longitude info that is a duplicate of the other stand-alone columns.


``` {r, echo=FALSE, warn=FALSE, message=FALSE}
library(tidyverse)
library(lubridate)
library(dplyr)
library(chron)
library(tidyverse)
library(ggplot2)
library(ggmap)
library(rnaturalearth)
library(rnaturalearthdata)
library(sf)
```


```{r}
colSums(is.na(df))
```

```{r, echo=FALSE}
df <- df %>% mutate(OCCUR_DATE = mdy(OCCUR_DATE))
head(df)
```

---

# Data Summary
``` {r}
summary(df)
```
# Extracting dimensions
From the existing data, I will extract additional dimensions to help with further analysis of the data.

```{r}
df$OCCUR_TIME_FMT <- strptime(df$OCCUR_TIME, "%H:%M:%S", tz="EST")
df$HOUR <- as.numeric(substr(df$OCCUR_TIME, start=1, stop=2))
df$MONTH <- as.numeric(format(as.Date(df$OCCUR_DATE, format="%Y-%m-%d"), "%m"))
df$YEAR <- as.numeric(format(as.Date(df$OCCUR_DATE, format="%Y-%m-%d"), "%Y"))
```

# Answering questions
- How often do gun crimes result in murder?
- Is there any correlation between murder and date or time?
- Is there any correlation between non-murder gun crimes and date or time?
- How does location affect gun crime? 
-- Which area has the most gun crime in general?
-- Which has the most non-murder gun crime?
-- Which has the most murder-flagged gun crime?

```{r}
# how often do gun crimes result in murder?
murder_vs_non <- df %>%
  group_by(YEAR, STATISTICAL_MURDER_FLAG) %>%
  summarise(TOTAL=n())
show(murder_vs_non)
ggplot(murder_vs_non, aes(x=YEAR, y=TOTAL, fill=STATISTICAL_MURDER_FLAG)) +  
  geom_bar(position="dodge", stat="identity") +
  scale_fill_discrete(labels=c('Not Murder', 'Murder')) + 
  ggtitle("Gun Crime by Murder Flag over Time") +
  labs(fill="Crime Type")
```

```{r}
ratio_df <- murder_vs_non %>% pivot_wider(names_from=STATISTICAL_MURDER_FLAG, values_from=TOTAL)
ratio_df$TOTAL <- (ratio_df$true + ratio_df$false)
ratio_df$RATIO <- (ratio_df$true / ratio_df$TOTAL) * 100
show(ratio_df)

ggplot(ratio_df) + 
  geom_line(aes(x=YEAR, y=RATIO), color = 'blue') +
  ggtitle("Gun Crime to Murder Ratio")
```


```{r}
gb <- df %>%
  group_by(MONTH, HOUR) %>%
  summarize(count_by_month = n())
piv_mtrx <- gb %>% pivot_wider(names_from=MONTH, values_from=count_by_month)
show(gb)
```


```{r}
ggplot(
  gb, 
  mapping=aes(
    x=MONTH, 
    y=HOUR,
    fill=count_by_month
    )
  ) + geom_tile()+
  scale_fill_gradient(low='white', high='red') +
  ggtitle("Count of Incidents by Month and Hour of the Day")
```


```{r, figures-side, fig.show="hold", out.width="50%"}

by_hour <- df %>% group_by(MONTH) %>%
  summarize(count_by_month = n())

by_month <- df %>% group_by(MONTH) %>%
  summarize(count_by_month = n())

#show(by_hour)

barplot(table(df$HOUR))
barplot(table(df$MONTH))

```
```{r}
ggmap::register_google()
```


```{r}
ymax = max(df$Latitude)
ymin = min(df$Latitude)
xmax = max(df$Longitude) # these are backwards because negatives
xmin = min(df$Longitude)
murders <- filter(df, STATISTICAL_MURDER_FLAG == "true")
murders <- murders %>% select('Latitude', 'Longitude')
non_murders <- non_murders %>% select('Latitude', 'Longitude')
non_murders <- filter(df, STATISTICAL_MURDER_FLAG == "false")
world <- ne_countries(scale = "medium", returnclass = "sf")
ggplot(data = world) + 
  geom_sf() +
  geom_point(data=murders, aes(x=Longitude, y=Latitude), color='red') +
  coord_sf(xlim = c(xmin, xmax), ylim = c(ymin, ymax), expand=FALSE) +
  ggtitle("Murders overlay Map")

ggplot(data = world) + 
  geom_sf() +
  geom_point(data=non_murders, aes(x=Longitude, y=Latitude), color='blue') +
  coord_sf(xlim = c(xmin, xmax), ylim = c(ymin, ymax), expand=FALSE) +
  ggtitle("Non-Murder Overlay Map")

```

```{r}
age_grps <- filter(df, !is.null(PERP_AGE_GROUP))
age_grps <- filter(df, 
                   !PERP_AGE_GROUP == "224" &
                   !PERP_AGE_GROUP == "940" &
                   !PERP_AGE_GROUP == "1020" &
                   !PERP_AGE_GROUP == ""
                   )

age_grps <- age_grps %>%
  group_by(PERP_AGE_GROUP, VIC_AGE_GROUP) %>%
  summarise(COUNT=n())

show(age_grps)

ggplot(
  age_grps, 
  mapping=aes(
    x=VIC_AGE_GROUP, 
    y=PERP_AGE_GROUP,
    fill=COUNT
    )
  ) + geom_tile() +
  scale_fill_gradient(low='white', high='red') +
  xlab("Victim Age Group") +
  ylab("Perpretrator Age Group") +
  ggtitle("Perpetration/Victim Age Group Interaction Matrix")
```

